Experiences with Planning for Natural Language Generation

### 2009-05-04 trunk/Papers/ci-spark/manuscript.tex
strengthened intro; updated graffle picture for PDF

\title{Experiences with Planning for Natural Language Generation}
\author{\textbf{Alexander Koller} \\
Exzellenzcluster / FR 4.7 Computerlinguistik \\
Saarland University \\
Postfach 151150 \\
66041 Saarbrücken \\
Germany \\
Phone: +49 681 302 70040 \\
Fax: +49 681 302 4351 \\
E-mail: \url{koller@mmci.uni-saarland.de}
\and
\textbf{Ronald P. A. Petrick} \\
School of Informatics \\
University of Edinburgh \\
Informatics Forum, 10 Crichton Street \\
Edinburgh \ EH8 9AB \\
Scotland, United Kingdom \\
Phone: +44 131 650 4426 \\
Fax: +44 131 650 6899 \\
E-mail: \url{rpetrick@inf.ed.ac.uk}}

@@

@@

Natural language generation (NLG; \citealp{reiter00building}) is one of the
major subfields of natural language processing, concerned with computing
natural language sentences or texts that convey a given piece of
information to the user. Intuitively, this task can be viewed as a problem
involving actions, beliefs, and goals: an agent communicating with another
agent tries to change the mental state of the hearer by applying actions
which correspond to the utterance of words or sentences.
This characterization of NLG suggests obvious parallels to automated
planning---a view which has a long tradition in NLG.

Early approaches to generation as planning
\citep{perrault80,appelt:planning,hovy88,young94dpocl} focused primarily on
planning high-level structures, such as speech acts and discourse
relations.  While these approaches suffered from the inefficiency of the
planners that were available at the time, there has recently been a
resurgence of interest in applying modern planning techniques to NLG.
These newer papers revisit speech act planning based on the more expressive
planning formalisms available today \citep{Steedman-Petrick:07,benotti08b}
and exploit the improved efficiency of modern planners for an integrated
model of sentence planning and surface realization \citep{KolSto07}.
Problems in NLG are often computationally challenging (see, e.g.,
\citep{KolStr02}), and these new approaches combine an interest in planning
as a modelling tool for natural language, with a hope of improved
efficiency by exploiting modern approaches and algorithms.

The focus of this paper is twofold. First, we present two novel planning
domains that arise in the context of natural language generation: the
sentence generation domain and the GIVE domain. In the sentence generation
task, the goal is to generate a single sentence that expresses a given
meaning. \citet{KolSto07} cast this domain as a planning problem, where the
plan encodes the necessary sentence and the actions correspond to uttering
individual words.  In the GIVE domain (``Generating Instructions in Virtual
Environments''), we describe a new shared task that was recently posed as a
challenge for the NLG community \citep{alexander07:_shared_task_propos}.
GIVE uses planning as part of an NLG system that generates natural-language
instructions to guide a user in performing a given task in a virtual
environment.

Second, we discuss some of our experiences using off-the-shelf
planners in our two NLG-inspired domains. In particular, we explore the
efficiency of FF \citep{HoffmannNebel01} and SGPLAN
\citep{hsu06:_new_featur_in_sgplan_for}: two planners that are readily
available, support an expressive subset of the Planning Domain Definition
Language (PDDL; McDermott et al.~\citeyear{PDDL}) for encoding domains, and
have been successful on traditional benchmarks and problems from the
International Planning Competition (IPC)\footnote{See
 \url{http://ipc.icaps-conference.org/} for information about past
 editions of the IPC.}.
We test these planners on a range of problem instances in our planning
domains, and compare our results against an ad-hoc Java implementation of
GraphPlan \citep{Blum1997}. Overall, our findings are mixed. On the one
hand, we demonstrate that modern planners can easily handle some of the
moderately-sized search problems that arise in NLG. In certain cases, the
planners we tested are more efficient than specialized NLG algorithms
applied to the same problem. On the other hand, these same planners spend
tremendous amounts of time on preprocessing. For instance, in the sentence
generation domain, FF spends 90\% of its runtime grounding out literals and
actions, and ends up being slower than our version of GraphPlan that avoids
grounding. For domains like ours, which are dominated by the number of
actions and the universe size, rather than the combinatorics of the search
problem, this observation suggests that the overall runtime of a planner
could be improved by grounding more selectively. This is particularly true
for large problem instances, which still remain a challenge for the
planners we tested, and where the effects of the grounding approach become
quite pronounced. We therefore offer our NLG domains as challenges for the
planning community.

\section{Planning in NLG}
\label{sec:domains}

We begin by presenting our two planning domains: the sentence generation
domain and the instruction giving domain.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.75\columnwidth]{pic-derivation-2}
  \caption{Derivation of ``The white rabbit sleeps.''}
  \label{fig:white-rabbit-sleeps-deriv}
\end{figure}

>>>>

\title{Experiences with Planning for Natural Language Generation}
\author{\textbf{Alexander Koller} \\
Exzellenzcluster / FR 4.7 Computerlinguistik \\
Saarland University \\
Postfach 151150 \\
66041 Saarbr\"ucken \\
Germany \\
Phone: +49 681 302 70040 \\
Fax: +49 681 302 4351 \\
E-mail: \url{koller@mmci.uni-saarland.de}
\and
\textbf{Ronald P. A. Petrick} \\
School of Informatics \\
University of Edinburgh \\
Informatics Forum, 10 Crichton Street \\
Edinburgh \ EH8 9AB \\
Scotland, United Kingdom \\
Phone: +44 131 650 4426 \\
Fax: +44 131 650 6899 \\
E-mail: \url{rpetrick@inf.ed.ac.uk}}

@@

\newcommand{\todo}[1]{\textbf{(#1)}}

@@

Natural language generation (NLG; \citealp{reiter00building}) is one
of the major subfields of natural language processing, concerned with
computing natural language sentences or texts that convey a given
piece of information to the user. This task can be viewed as a problem
of achieving a (communicative) goal by successively applying
(communicative) actions, and thus has clear intuitive parallels to
automated planning.

This view of generation as planning has a long tradition in NLG
\citep{perrault80,appelt:planning,hovy88,young94dpocl}. Despite their
theoretical attractiveness, these early approaches never gained
mainstream use in applied systems, in part because of the inefficiency
of the planners that were available at the time. However, the recent
improvements in planner efficiency and expressiveness has sparked a
renewed interest in applying modern planning techniques to NLG---both
on the level of speech act planning
\citep{Steedman-Petrick:07,benotti08b} and and on the level of
sentence generation \citep{KolSto07}. The goal of this paper is to
determine, by looking at some representative generation problems,
whether planning has advanced to the point that this enterprise can be
successful this time.

The focus of this paper is twofold. First, we present two generation
problems that have recently been cast as planning problems: the
sentence generation task and the GIVE task. In the sentence generation
task, the goal is to generate a single sentence that expresses a given
meaning. \citet{KolSto07} cast this task as a planning problem, where
the plan encodes the necessary sentence and the actions correspond to
uttering individual words.  In the GIVE domain (``Generating
Instructions in Virtual Environments''), we describe a new shared task
that was recently posed as a challenge for the NLG community
\citep{ByrKolStrCasDalMooObe09}.  GIVE uses planning as part of an NLG
system that generates natural-language instructions to guide a user in
performing a given task in a virtual environment.

Second, we discuss some of our experiences using off-the-shelf
planners in our two NLG-inspired domains. In particular, we explore
the efficiency of FF \citep{HoffmannNebel01} and SGPLAN
\citep{hsu06:_new_featur_in_sgplan_for}: two planners that are readily
available, support an expressive subset of the Planning Domain
Definition Language (PDDL; McDermott et al.~\citeyear{PDDL}) for
encoding domains, and have been successful on traditional benchmarks
and problems from the International Planning Competition
(IPC)\footnote{See \url{http://ipc.icaps-conference.org/} for
  information about past editions of the IPC.}.  We test these
planners on a range of problem instances in our planning domains, and
compare our results against an ad-hoc Java implementation of GraphPlan
\citep{Blum1997}. Overall, our findings are mixed. On the one hand, we
demonstrate that modern planners can easily handle the \emph{search}
problems that arise in NLG on realistic inputs, which is promising
given that the sentence generation task, at least, is NP-complete
\citep{KolStr02}. On the other hand, these same planners spend
tremendous amounts of time on preprocessing. For instance, in the
sentence generation domain, FF spends 90\% of its runtime grounding
out literals and actions, and ends up being slower than our version of
GraphPlan that avoids grounding. For domains like ours, which are
dominated by the number of actions and the universe size, rather than
the combinatorics of the search problem, this observation suggests
that the overall runtime of a planner could be improved by grounding
more selectively. This is particularly true for large problem
instances, which still remain a challenge for the planners we tested,
and where the effects of the grounding approach become quite
pronounced. We conclude that planning \emph{algorithms} have advanced
to the point that generation as planning is now a realistic endeavour,
but the concrete \emph{implementations} of these algorithms must be
improved to be useful in this application, and offer our NLG domains
as challenges for the planning community.

This paper is structured as follows. We begin by briefly reviewing
some of the literature on NLG as planning
(Section~\ref{sec:nlg-as-planning}. Then we define introduce the
planning problems arising in two particular NLG tasks in more detail
in Section~\ref{sec:domains}. We report on some experiments with these
planning problems in Section~\ref{sec:experiments}, and discuss the
results in Section~\ref{sec:discussion}. Section~\ref{sec:conclusion}
concludes.

\section{NLG as Planning} \label{sec:nlg-as-planning}

The task of generating natural language from semantic representations
is typically split up into two parts: the \emph{domain planning} task,
which selects the information that is to be conveyed and structures it
into sentence-sized chunks, and the \emph{sentence generation} tasks,
which then translates each of these chunks into natural language
sentences. The sentence generation task is often split up into two
parts of its own---the \emph{sentence planning} task, which enriches
the input by e.g.\ determining how to refer to objects and selects
some lexical material, and the \emph{surface realization} task, which
maps the enriched meaning representation into a sentence using a
grammar. The chain of domain planning, sentence planning, and surface
realization is sometimes called the ``NLG pipeline''
\cite{reiter00building}.

Viewing generation as a planning problem has a long tradition in the
NLG literature, especially in discourse planning. \todo{Sketch
  Perrault et al here. Say something very brief about Appelt, Hovy,
  Moore. Say why all this didn't really work.}

Most recently, interest in generation as planning has been revived for
NLG tasks on various levels of the NLG pipeline. Below, we will look
in some detail at \cite{KolSto07}'s recasting of sentence generation
as planning; furthermore, \cite{Steedman-Petrick:07} and
\cite{benotti08b} have applied planning to \todo{something}. Notice
that although all of these approaches encode ``communicative actions''
as planning operators, they still solve fundamentally different
problems. For instance, Perrault et al.\ tackled discourse planning
with planning operators that encode speech acts, each of which
represents the meaning of a whole sentence and models the contribution
of this sentence to the intentional structure of the entire
discourse. By contrast, Koller and Stone solve a sentence generation
problem using planning operators that encode utterances of single
words and model the contribution of each word to the grammatical
structure of the sentence. Whether these different levels of ``NLG as
planning'' could be combined into a single planning problem, and
whether this would make computational sense, is an open question.

Problems in NLG at all levels of description can become
computationally complex very easily. For instance, \citet{KolStr02}b
showed that even surface realization is NP-complete under reasonable
assumptions. This makes it necessary to investigate methods for
guiding the search, which makes modern planning techniques a promising
approach to NLG. The point of this paper is to investigate whether
existing planners fulfill this promise of efficiency for the
application to NLG.

\section{Two NLG Tasks}
\label{sec:domains}

We will now focus the discussion towards two specific NLG problems:
sentence generation in the sense of \cite{KolSto07}, and the
generation of instructions in virtual environments
\cite{ByrKolStrCasDalMooObe09}.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width=0.75\columnwidth]{pic-derivation}
  \caption{Derivation of ``The white rabbit sleeps.''}
  \label{fig:white-rabbit-sleeps-deriv}
\end{figure}
